Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-05-16 writeups/Su_weak_tides.tex
investigated scaling relation for tCE2 probability at small eta_sync

% \section{Separatrix Crossing Probability with Weak Tidal Friction
% }\label{app:sep_crossing_dynamics}

% In this appendix, we analytically calculate $\Delta K_{\pm}$ for use in
% Eq.~\eqref{eq:def_pc_weaktide} to calculate the probabilities of the two
% possible transitions upon separatrix encounter. We first rewrite the full
% equations of motion for the planet's spin including weak tidal friction in
% component form:
% \begin{align}
%     \rd{\theta}{t} &= g\sin I \sin \phi -
%         \frac{1}{t_{\rm a}} \sin \theta,\label{eq:ds_fullq}\\
%     \rd{\phi}{t} &= -\alpha\cos\theta
%         - g\p{\cos I + \sin I \cot \theta \cos \phi}\label{eq:ds_fullphi},\\
%     \rd{\Omega_{\rm s}}{t}
%         &= \frac{1}{t_{\rm a}} \s{2n \cos \theta
%             - g\Omega_{\rm s}\p{1 + \cos^2\theta}}\label{eq:ds_fulls}.
% \end{align}

% Note that Eq.~\eqref{eq:def_dK_weaktide} can be rewritten
% \begin{align}
%     \Delta K_{\pm} &= \oint_{\mathcal{C}_{\pm}} \rd{H}{t}
%         - \rd{H_{\rm sep}}{t}\;\mathrm{d}t
%         \nonumber\\
%         &= \oint_{\mathcal{C}_{\pm}}
%            \p{\rd{(\cos\theta)}{t}}_{\rm tide}
%             + \frac{\dot{\Omega}_{\rm s}}{\dot{\phi}}
%             \p{\pd{H}{s} - \pd{H_{\rm sep}}{s}}\;\mathrm{d}\phi
%                 \label{eq:app_dhpm}.
% \end{align}

% We then evaluate $\Delta K_{\pm}$ by integrating along the two legs of the
% separatrix $\mathcal{C}_{\pm}$ (see Fig.~\ref{fig:1contours}). Note that we must
% use the value of $\eta$ at the moment of separatrix encounter, as the evolution
% of $\Omega_{\rm s}$ changes the spin-orbit precession frequency $\alpha$
% \begin{align}
%     \Delta K_{\pm} ={}& \oint_{\mathcal{C}_{\pm}}
%         \frac{1}{t_{\rm s}}\sin^2\theta
%             \p{\frac{2n}{\Omega_{\rm s}} - \cos\theta}
%          +
%             \s{\frac{\alpha}{\Omega_{\rm s}}
%                 \frac{\cos^2\theta}{2} + g\frac{\Omega_{\rm
%             c}}{2\Omega_{\rm s}^2}\cos^2 I}\rd{\Omega_{\rm
%             s}}{t}\rd{t}{\phi}\;\mathrm{d}\phi\\
%      \approx{}& \oint_{\mathcal{C}_{\pm}}
%         \frac{1}{t_{\rm s}}\sin^2\theta
%             \p{\frac{2n}{\Omega_{\rm s}} - \cos\theta}
%         + \frac{1}{t_{\rm s}}\frac{n}{\Omega_{\rm s}\eta}
%             \s{\p{\cos \theta}_{\mathcal{C}_{\pm}} - \frac{\Omega_{\rm s}}{2n}}
%             \s{2 \cos I \pm \sqrt{2 \sin I\p{1 - \cos \phi} / \eta}}
%             \;\mathrm{d}\phi.\label{eq:app_deltaK}
%     t_{\rm s} \Delta K_{\pm} \approx{}&
%         -2\cos I\p{\pm 2\pi \eta \cos I + 8\sqrt{\eta \sin I}}
%         \pm 2\pi \frac{\Omega_{\rm s}}{n}\cos I
%         - 8\eta \cos I \sqrt{\sin I / \eta}
%             + \frac{4\Omega_{\rm s}}{n}\sqrt{\sin I/\eta}\nonumber\\
%         &+ \frac{2n}{\Omega_{\rm s}}\p{\mp 2\pi\p{1 - 2\eta \sin I}
%             + 16\cos I \eta^{3/2}\sqrt{\sin I}}
%             + 8\sqrt{\eta \sin I}
%             \pm 2 \pi \eta \cos I
%             - \frac{64}{3} \p{\eta \sin I}^{3/2}.\label{eq:app_deltaK}
% \end{align}
% At this point, since $\eta$ often is not small, it is necessary to integrate
% along the numerically computed $\mathcal{C}_{\pm}$ (instead of the approximate
% solution given by Eq.~\ref{eq:sep_theta}) to obtain sufficiently accurate values
% for $\Delta K_{\pm}$. These numerically integrated values are used in
% Eq.~\eqref{eq:def_pc_weaktide} to semi-analytically compute the tCE2
% probabilities (blue curves) in
% Figs.~\ref{fig:pc_fits_0_06}--\ref{fig:pc_fits_0_20}.

>>>>

\section{Approximate TCE2 Probability for Small $\eta_{\rm sync}$
}\label{app:sep_crossing_dynamics}

In this appendix, we seek a tentative analytic understanding for the
probability of convergence to tCE2 when $\eta_{\rm sync}$ is small, i.e.\ the
left extremes of Figs.~\ref{fig:probs20} and~\ref{fig:probs5}. In this regime,
following the discussions in Sections~\ref{ss:toy_outcomes}
and~\ref{ss:phop_weaktide}, we understand that initial conditions in zone I
always converge to tCE1, initial conditions in zone II always converge to tCE2,
and initial conditions in zone III experience separatrix encounter and
probabilistically converge to either one of the tCE\@. To further proceed, we
will assume an isotropic distribution of initial spin orientations; different
distributions again will only change the quantitative but not qualitative
character of the discussion.

To express the contribution of initial conditions in zone II to the total tCE2
probability (green shaded region in Figs.~\ref{fig:probs20}
and~\ref{fig:probs5}), we need only to know the probability an arbitrary
initial condition is in zone II\@, which can be obtained by the phase space
area of zone II\@. An exact analytic form for this area is known
(\citealp{ward2004I}, Paper I), but an accurate approximation can be obtained
using Eq.~\eqref{eq:sep_theta} since $\eta_{\rm i} \ll 1$:
\begin{align}
    \frac{A_{\rm II}}{4\pi} = \frac{4}{\pi}\sqrt{\eta_{\rm i} \sin I}.
\end{align}
where $A_{\rm II}$ is the phase space area (in $\cos \theta$ and $\phi$
coordinates) of zone II\@.

To express the contribution of initial conditions in zone III to the totac tCE2
probability (blue shaded region in Figs.~\ref{fig:probs20}
and~\ref{fig:probs5}), we need both the area of zone III and the distribution of
outcomes upon separatrix encounter. The area of zone III can again be estimated
quite simply using Eq.~\eqref{eq:sep_theta}, and we obtain
\begin{align}
    \frac{A_{\rm III}}{4\pi}
        = \frac{1 + \eta_{\rm i} \cos I}{2} -
            \frac{2}{\pi}\sqrt{\eta_{\rm i} \sin I}.
\end{align}
where $A_{\rm III}$ is the phase space area of zone III\@. Next, we need to
understand the outcomes of the separatrix encounters that initial conditions in
zone III experience.. We analytically calculate $\Delta K_{\pm}$
(Eq.~\ref{eq:def_dK_weaktide}) for use in Eq.~\eqref{eq:def_pc_weaktide} to
obtain the probabilities of the outcomes of separatrix encounter. We first
rewrite Eq.~\eqref{eq:def_dK_weaktide} as:
\begin{align}
    \Delta K_{\pm} &= \oint_{\mathcal{C}_{\pm}} \rd{H}{t}
        - \rd{H_{\rm sep}}{t}\;\mathrm{d}t
        \nonumber\\
        &= \oint_{\mathcal{C}_{\pm}}
           \p{\rd{(\cos\theta)}{t}}_{\rm tide}
            + \frac{\dot{\Omega}_{\rm s}}{\dot{\phi}}
            \p{\pd{H}{s} - \pd{H_{\rm sep}}{s}}\;\mathrm{d}\phi
                \label{eq:app_dhpm}.
\end{align}
Then, using the full equations of motion for the planet's spin including weak
tidal friction in component form, given by
Eqs.~(\ref{eq:ds_fullq}--\ref{eq:ds_fulls}), we can evaluate $\Delta K_{\pm}$ by
integrating along the two legs of the separatrix $\mathcal{C}_{\pm}$ (see
Fig.~\ref{fig:1contours}). Note that we must use the value of $\eta$ at the
moment of separatrix encounter, which we denote $\eta_{\rm cross}$, as the
evolution of $\Omega_{\rm s}$ changes the spin-orbit precession frequency
$\alpha$ and thus $\eta$ itself:
\begin{align}
    % \Delta K_{\pm} ={}& \oint_{\mathcal{C}_{\pm}}
    %     \frac{1}{t_{\rm s}}\sin^2\theta
    %         \p{\frac{2n}{\Omega_{\rm s}} - \cos\theta}
    %      +
    %         \s{\frac{\alpha}{\Omega_{\rm s}}
    %             \frac{\cos^2\theta}{2} + g\frac{\Omega_{\rm
    %         c}}{2\Omega_{\rm s}^2}\cos^2 I}\rd{\Omega_{\rm
    %         s}}{t}\rd{t}{\phi}\;\mathrm{d}\phi\nonumber\\
    %  \approx{}& \oint_{\mathcal{C}_{\pm}}
    %     \frac{1}{t_{\rm s}}\sin^2\theta
    %         \p{\frac{2n}{\Omega_{\rm s}} - \cos\theta}
    %     + \frac{1}{t_{\rm s}}\frac{n}{\Omega_{\rm s}\eta_{\rm cross}}
    %         \s{\p{\cos \theta}_{\mathcal{C}_{\pm}} - \frac{\Omega_{\rm s}}{2n}}
    %         \s{2 \cos I \pm \sqrt{2 \sin I\p{1 - \cos \phi} / \eta_{\rm cross}}}
    %         \;\mathrm{d}\phi,\\
    t_{\rm s} \Delta K_{\pm} \approx{}&
        \frac{\eta_{\rm cross}^2}{\eta_{\rm sync}}\s{
            -2\cos I\p{\pm 2\pi \eta_{\rm cross} \cos I
                + 8\sqrt{\eta_{\rm cross} \sin I}}
            \mp 4\pi \sin I
            - 8 \cos I \sqrt{\eta_{\rm cross}\sin I}
                + \frac{4\eta_{\rm sync}}{\eta_{\rm cross}}
                    \sqrt{\sin I/\eta_{\rm cross}}}\nonumber\\
        &+ \frac{2\eta_{\rm cross}}{\eta_{\rm sync}}
            \p{\mp 2\pi\p{1 - 2\eta_{\rm cross} \sin I}
            + 16\cos I \eta_{\rm cross}^{3/2}\sqrt{\sin I}}
            + 8\sqrt{\eta_{\rm cross} \sin I}
            \pm 2 \pi \eta_{\rm cross} \cos I
            - \frac{64}{3} \p{\eta_{\rm cross} \sin I}^{3/2}.
                \label{eq:app_deltaK}
\end{align}
The resulting $P_{\rm III \to II}$ obtained using this analytic $\Delta K_{\pm}$
in Eq.~\eqref{eq:def_pc_weaktide} is shown as the green dashed line in the top
panel of Fig.~\ref{fig:pc_fits_0_06}, where it can be seen that agreement is
reasonable for $\eta_{\rm cross} \lesssim 0.05$. For the purposes of this
section, we drop all but the leading order terms in both the numerator and
denominator of Eq.~\eqref{eq:def_pc_weaktide} and obtain:
% proof:
% from utils import *; I = np.radians(20); s_c = 0.01; s = np.linspace(1, 10, 100); top, bottom = get_ps_anal(I, s_c, s); plt.plot(s_c/s, (bottom + top) / bottom, 'k'); plt.plot(s_c/s, 6 * s_c / np.pi * np.sqrt(np.sin(I) / (s_c / s)), 'g')
\begin{equation}
    P_{\rm III \to II} \approx
        \frac{6\eta_{\rm sync}}{\pi} \sqrt{\frac{\sin I}{\eta_{\rm cross}}}.
\end{equation}
Finally, based on the bottom panel of Fig.~\ref{fig:pc_fits_0_06}, we make the
approximation that $\eta_{\rm cross}$ is uniformly distributed between
$\eta_{\rm i}$ and $\eta_{\rm sync}$. This approximation is mildly justifiable
as $\eta_{\rm cross}$ favors smaller values in the interval $\s{\eta_{\rm i},
\eta_{\rm sync}}$, but also extends beyond $\eta_{\rm sync}$ for $\cos
\theta_{\rm i}$ close to $-1$, so we hope these two effects cancel out to some
extent. Thus,
\begin{align}
    \ev{P_{\rm III \to II}}
        &\sim \frac{1}{\eta_{\rm sync} - \eta_{\rm i}}
            \int\limits_{\eta_{\rm i}}^{\eta_{\rm sync}}
            P_{\rm III \to II}\;\mathrm{d}\eta_{\rm cross}\nonumber\\
        &= \frac{12 \sqrt{\eta_{\rm sync}\sin I}}{\pi
            \p{1 + \sqrt{n / \Omega_{\rm s, i}}}}.
\end{align}

Finally, returning to the aggregate tCE2 probability, we can evaluate it as:
\begin{align}
    \text{tCE2 Probability}
        &= \frac{A_{\rm II}}{4\pi}
            + \frac{A_{\rm III}}{4\pi}\ev{P_{\rm III \to II}}\nonumber\\
        &\simeq
            \frac{4\sqrt{\eta _{\rm sync} \sin I}}{\pi}\p{
                \sqrt{n / \Omega_{\rm s, i}}
                + \frac{3}{2\p{1
                + \sqrt{n / \Omega_{\rm s, i}}}}}
            + \mathcal{O}\p{\eta_{\rm sync}}.\label{eq:app_tce2_p_tot}
\end{align}
